Software Product Line Analysis Techniques

### 2018-01-12 doc/quals/quals.tex
Abstract Interpretation - take 2

In \cite{Midtgaard:2015}, a systematic approach to lift abstract interpretation analyses to SPLs is presented. The paper provides a rigorous  formal treatment of abstract interpretation applied to a simple imperative language, with variability at the statement level. Formal proofs for the correctness of the lifting approach are also presented.

The lifting approach provides systematic liftings of the abstract domains, Galois connections, and corresponding analyses by the use of combinators. It is worth noting that this approach is systematic but not automated. Language-specific theorems need to be proven (3 of them have been proven in the paper), in addition to a language-agnostic framework theorem that has been proven as well.

The paper also presents some optimizations that would trade analysis precision for speed. The idea is to allow for variational abstraction, which would further abstract variability. Further optimization techniques involve sharing analysis intermediate values while keeping track of presence conditions (either in CNF/DNF formulas or BDDs). In addition, a technique for extracting data-flow equations (both the product-level equations and the lifted equations) is explained.

This work was evaluated on a Reaching Definitions analysis for Java programs. Three Java SPLs were used: GPL, Prevayler and BerkeleyDB. The speedup due to the different optimizations for the lifting approach (with respect to brute force) is exponential with respect to the number of SPL configurations.

>>>>

\newcommand{\configSpace}{\mathbb{K}}
\newcommand{\featSpace}{\mathbb{F}}
\newcommand{\lang}{\mathcal{L}}

In \cite{Midtgaard:2015}, a systematic approach to lift abstract interpretation analyses from single programs to SPLs is presented. The paper provides a rigorous  formal treatment of abstract interpretation applied to a simple imperative language, with variability at the statement level. Formal proofs of the soundness of the lifting approach are also presented.

Given a programming language $\lang$, starting with an abstract intepretation (collecting semantics, an abstract domain, a Galois connection connecting the two, and an analysis function), the paper presents the \emph{lift} combinator, which lifts each of those components to tuples of values, with each value corresponding to a product configuration in the configuration space $\configSpace$. If $\lang$ is extended with static conditionals (e.g., CPP \#if) into a variability-aware language  $\lang'$, and the semantics are respectively extended, the \emph{lift} combinator can take a product-level abstract interpretation analysis and generate a variability-aware analysis out of it.

The formulation presented lifts single values to tuples of values, each tuple $|\configSpace|$ in length. $|\configSpace|$ is exponential in the number of features in the feature space $\featSpace$, so implementing lifted analyses this way would be intractable. Caching and memoization of intermediate values can save some computational cycles, but still the overall computational complexity of the lifted analysis will grow exponentially with the number of features. Two optimizations are proposed in the paper: \emph{value sharing} and \emph{variational abstraction}.

\emph{Value sharing} attempts to eliminate redundancy within a lifted tuple by simply mapping individual values to presence conditions, instead of enumerating the values belonging to each of the configurations. This way if all configurations share the same values except for one, the tuple would only have two values with a presence condition each. The shorter the tuple, the faster the analysis since it will need to process less intermediate values. Presence conditions can be efficiently encoded and checked for satisfiability as CNF/DNF formulas or BDDs.

\emph{Variational abstraction} pushes abstraction up to the configuration space. Instead of lifting the exact product-level analysis up to the SPL, the lifted analysis can be abstracted to an over-approximation. Pretty much like all abstraction techniques, this trades accuracy for performance. \emph{Variational abstraction} is implemented the same as other abstraction stages, starting with an abstract domain and a Galois connection together with all the corresponding proofs. Abstraction can vary from minimal ($\alpha_\featSpace$ where no abstraction takes place) to maximal ($\alpha_\phi$ where collapses the configuration space into a single value). Since Galois connections compose (the composition of two Galois connections is a Galois connection), \emph{variational abstraction} fits naturally in the presented calculational abstract interpretation framework.

This lifting approach was evaluated on a single analysis (reaching definitions). Three Java SPLs (Graph PL (GPL), Prevayler and BerkeleyDB) were used. However, the evaluation numbers presented are not aggregated across the 3 codebases, but rather for only 4 source code methods (2 from GPL, 1 from Prevayler and 1 from BerkeleyDB). The configuration space of each of those methods varies between 8 and 106 configurations.

The performance (running time) of three SPL analysis techniques is measured: brute force (per product analysis), lifted analysis with sharing, and lifted analysis with maximum variational abstraction (abstracting configuration variability away). Comparing against brute force analysis, speedup due to lifting with maximum variational abstraction grows exponentially with respect with the number of configurations, and almost identical to the average analysis performance of a single configuration. The speedup due to lifting with sharing also grows exponentially, but at a slower rate. Generalizing from those performance numbers is risky though because of highly restricted sample space (a single analysis on only 4 source code methods).
